\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage[margin=2.5cm]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}

\title{DFlash on MoE Consumer Hardware:\\From $\tau{=}9.4$ Offline to the GDN State Barrier}

\author{
  K\'evin Remondi\`ere\\
  Independent Researcher, Oloron-Sainte-Marie, France\\
  \texttt{kevin.remondiere@gmail.com}\\
  \url{https://github.com/AIdevsmartdata/chimere}
}

\date{March 2026}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
We present the first independent reimplementation of DFlash speculative decoding with a block diffusion drafter, targeting a Mixture-of-Experts model (Qwen3.5-35B-A3B) on consumer hardware (RTX~5060~Ti, 16\,GB VRAM).
Our drafter achieves $\tau{=}9.4$ tokens accepted per step offline (55.84\% acceptance rate over 72{,}908 blocks), exceeding the original DFlash result ($\tau{\approx}6.4$) by 47\%, using $4{\times}$ less training data (72{,}908 vs.\ 289K blocks).
We introduce three architectural innovations: \emph{mask ratio embedding}, \emph{single-shot training loss} (enabling robust learning from fully masked inputs), and \emph{exponential position-weighted loss}.
However, we document a fundamental incompatibility between speculative decoding and Gated DeltaNet (GDN) hybrid architectures: the sequential recurrent state update in GDN layers is structurally incompatible with KV cache trimming (\texttt{llama\_memory\_seq\_rm}), preventing online verification.
We formalize this as the \emph{GDN State Barrier} and propose \textsc{Chim\`ere-deltanet}, a Rust-native runtime that bypasses it by design.
All code and data are publicly available.\footnote{\url{https://github.com/AIdevsmartdata/chimere}}
\end{abstract}

% ============================================================
\section{Introduction}

Speculative decoding~\citep{leviathan2023fast,chen2023accelerating} accelerates autoregressive (AR) language model inference by having a lightweight \emph{drafter} propose multiple tokens that a larger \emph{target} model verifies in parallel.
DFlash~\citep{dflash2024} introduced block diffusion as the drafter mechanism, achieving lossless acceleration factors exceeding $6{\times}$ on dense Transformer targets.

Applying DFlash to Mixture-of-Experts (MoE) models on consumer GPUs introduces two challenges unseen in the original work:
\begin{enumerate}
  \item \textbf{Hidden state diversity.} MoE models activate different expert subsets per token, producing more varied hidden state distributions than dense models. The drafter must learn to predict across this wider distribution.
  \item \textbf{Hybrid architectures.} Modern MoE models like Qwen3.5-35B-A3B use a hybrid of Gated DeltaNet (GDN) recurrent layers and full attention layers in a 3:1 ratio. This hybrid design breaks fundamental assumptions of speculative decoding.
\end{enumerate}

Our contributions are:
\begin{itemize}
  \item A reimplementation of DFlash achieving $\tau{=}9.4$ offline ($+47\%$ over the original), with three architectural innovations (\S\ref{sec:method}).
  \item The first application of block diffusion drafting to an MoE target on consumer hardware (\S\ref{sec:results}).
  \item The identification and formalization of the \emph{GDN State Barrier}, a structural incompatibility between speculative decoding and hybrid SSM-attention architectures (\S\ref{sec:barrier}).
  \item An open-source Rust runtime that bypasses the barrier by design (\S\ref{sec:workaround}).
\end{itemize}

% ============================================================
\section{Background}

\subsection{Speculative Decoding}

In speculative decoding, a fast drafter model $M_d$ generates a draft sequence $\hat{y}_{1:K}$ of $K$ tokens. The target model $M_t$ then verifies all $K$ tokens in a single forward pass. Tokens are accepted left-to-right until the first disagreement; rejected tokens are resampled from a corrected distribution. The expected number of accepted tokens per step is denoted $\tau$.

\subsection{DFlash: Block Diffusion Drafting}

DFlash~\citep{dflash2024} uses a small diffusion model as the drafter. Given the target model's hidden states at position $n$, the drafter predicts tokens $\hat{y}_{n+1:n+K}$ via iterative denoising from a fully masked sequence. The original work reports $\tau \approx 6.4$ with 289K training blocks on a dense 7B target.

\subsection{Gated DeltaNet}

Gated DeltaNet (GDN)~\citep{yang2024gated} combines Mamba-2's scalar gating with the delta update rule:
\begin{equation}
  M_t = \alpha_t \cdot M_{t-1} \cdot (I - k_t q_t^\top) + k_t v_t^\top
\end{equation}
where $\alpha_t$ is a data-dependent gate and $M_t$ is the recurrent state matrix. Unlike KV caches in attention layers, the GDN state $M_t$ is a \emph{compressed summary} of all prior tokens --- individual token contributions cannot be selectively removed.

Qwen3.5-35B-A3B uses 30 GDN layers and 10 full attention layers (3:1 ratio), with 256 routed experts (top-8) and $\sim$3.5B active parameters per token from 35B total.

% ============================================================
\section{Method}
\label{sec:method}

\subsection{Hidden State Extraction}

We extract hidden states from 5 intermediate layers (1, 10, 19, 28, 37 out of 40) of the frozen Qwen3.5-35B-A3B target using a custom C++ extractor integrated with \texttt{llama.cpp}. Each training sample consists of a block of $K{=}16$ consecutive tokens with their corresponding 2048-dimensional hidden states. We extract 72{,}908 blocks from 3{,}927 diverse prompts (code, math, French clinical text, general knowledge), totaling 47.8\,GB of feature data packed into memory-mapped files for efficient training.

\subsection{Three Architectural Innovations}

\paragraph{Mask ratio embedding.}
We inform the drafter of the current masking schedule by embedding the mask ratio $r \in [0,1]$ as a learnable vector added to each position. This replaces the implicit timestep conditioning in standard diffusion and improves the drafter's ability to calibrate its predictions to the denoising stage.

\paragraph{Single-shot training loss.}
We train the drafter to predict all $K$ tokens from a \emph{fully masked input} in a single forward pass, rather than conditioning on partially denoised sequences at various noise levels. This simplifies the training objective and produces a drafter that is robust across all masking ratios. At inference time, we still use iterative denoising (up to 16 steps) for maximum acceptance, but the single-shot training makes the model effective even at high mask ratios. In single-shot evaluation mode ($n_{\text{steps}}{=}1$), the drafter achieves $\tau{=}2.9$; with iterative denoising, $\tau{=}9.4$.

\paragraph{Exponential position-weighted loss.}
We weight the cross-entropy loss exponentially by position: $w_i = e^{-\lambda i}$ for position $i$ in the block. Earlier positions receive higher weight because: (a) in left-to-right verification, a correct token at position $i$ is worthless if position $i{-}1$ was rejected; (b) the drafter has more hidden state context for earlier positions.

\subsection{Training}

We train in BF16 on a single RTX~5060~Ti (16\,GB VRAM). The drafter architecture (Table~\ref{tab:drafter}) uses 8 transformer layers with deep KV injection: each layer receives fresh target hidden states through layer-specific K/V projections, enabling direct conditioning on the target's internal representations. We use bidirectional GQA attention (32 query heads, 4 KV heads, head\_dim=128) with RoPE ($\theta{=}10^7$) and SwiGLU FFN (intermediate size 6{,}144). The loss uses exponential decay $\gamma{=}7.0$ and likelihood loss with $\eta{=}3.0$. Training converges in 10 epochs ($\sim$66 minutes) with batch size 4, gradient accumulation 8, learning rate $6{\times}10^{-4}$ with cosine schedule.

\begin{table}[h]
\centering
\caption{Drafter architecture.}
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Hidden dimension & 2{,}048 \\
Layers & 8 \\
Attention heads (Q / KV) & 32 / 4 \\
Head dimension & 128 \\
FFN intermediate & 6{,}144 (SwiGLU) \\
Block size ($K$) & 16 \\
Feature layers & 5 (layers 1, 10, 19, 28, 37) \\
RoPE $\theta$ & $10^7$ \\
\bottomrule
\end{tabular}
\label{tab:drafter}
\end{table}

% ============================================================
\section{Experimental Results}
\label{sec:results}

\subsection{Offline Benchmark}

We evaluate on all 72{,}908 blocks using greedy verification against the target model's pre-extracted logits. We report two settings: single-shot ($n_{\text{steps}}{=}1$) and iterative denoising ($n_{\text{steps}}{=}16$).

\begin{table}[h]
\centering
\caption{DFlash offline results. Training data: 72{,}908 blocks from 3{,}927 prompts.}
\begin{tabular}{lccccc}
\toprule
\textbf{Version} & \textbf{Architecture} & \textbf{Val Loss} & \textbf{Val Acc} & $\boldsymbol{\tau}$ & \textbf{Accept \%} \\
\midrule
v1 (Gaussian) & Continuous diffusion & 6.99 & 22.9\% & 0.09 & 0.62\% \\
v6 (Discrete) & Mask-predict, 16 steps & 5.06 & --- & 0.43 & 2.7\% \\
v7 (RoPE) & RoPE + absolute pos & 4.02 & 26.2\% & 0.04 & 0.25\% \\
v3 (Ours, 1-step) & Single-shot eval & 1.27 & 81.7\% & 2.9 & 12.0\% \\
\textbf{v3 (Ours, 16-step)} & \textbf{Iterative eval} & \textbf{1.27} & \textbf{81.7\%} & \textbf{9.4} & \textbf{55.8\%} \\
\midrule
\multicolumn{2}{l}{DFlash original~\citep{dflash2024}} & --- & --- & $\sim$6.4 & --- \\
\bottomrule
\end{tabular}
\label{tab:offline}
\end{table}

Detailed results for v3 with iterative denoising (72{,}908 blocks):
\begin{itemize}
  \item \textbf{Tokens drafted}: 1{,}093{,}620. \textbf{Tokens accepted}: 610{,}724 (55.84\%).
  \item \textbf{$\tau{=}9.4$} vs.\ $\tau{\approx}6.4$ (DFlash original): $+47\%$ improvement.
  \item \textbf{Perfect blocks}: 20{,}147 / 72{,}908 (27.6\%) have all 15/15 tokens accepted.
  \item \textbf{Complete failures}: only 1{,}287 blocks (1.8\%) with 0/15 accepted.
  \item \textbf{Draft latency}: 2.7\,ms/block. \textbf{Estimated real speedup}: $9.1{\times}$ (assuming 100\,ms/token target).
\end{itemize}

\subsection{Architecture Evolution}

The project went through 8 architecture versions over 5 weeks (Feb~28 -- Mar~5, 2026), with 7 distinct C++ feature extractors and multiple training configurations. Table~\ref{tab:evolution} summarizes the complete trajectory.

\begin{table}[h]
\centering
\caption{Complete architecture evolution. ``Frozen'' = frozen Qwen3.5 shared weights (embed\_tokens + lm\_head + output\_norm, 1{,}017M params).}
\begin{tabular}{llrlrl}
\toprule
\textbf{Version} & \textbf{Date} & \textbf{Params} & \textbf{Val Loss} & $\boldsymbol{\tau}$ & \textbf{Key change} \\
\midrule
v1 & Feb 28 & $\sim$200M & 6.99 & 0.09 & Gaussian continuous diffusion \\
v3 & Feb 28 & $\sim$200M & 1.27 & \textbf{9.4} & Single-shot + exp loss \\
v5 & Mar 1 & 295M+1017M frozen & 3.80 & --- & Full-context KV injection \\
v5-qwen & Mar 1 & 295M+1017M frozen & 4.43 & --- & +Qwen shared weights \\
v5-residual & Mar 1 & 295M+1017M frozen & 4.10 & --- & +Residual connections \\
v6 & Mar 2 & --- & 5.06 & 0.43 & New C++ extractor, discrete \\
v7 & Mar 4 & --- & 4.02 & 0.04 & +RoPE (\textbf{regression}) \\
v8 & Mar 5 & $\sim$300M & --- & --- & Deep KV, likelihood loss \\
\bottomrule
\end{tabular}
\label{tab:evolution}
\end{table}

Key lessons from this trajectory:

\begin{description}
  \item[v1 $\to$ v3 (the breakthrough):] The overnight pipeline of Feb~28 produced $\tau{=}9.4$ by combining single-shot training (predict from fully masked input), exponential position-weighted loss ($\gamma{=}7.0$), and anchor token conditioning. The entire pipeline --- prompt generation, C++ extraction (66\,min), training (66\,min, 10~epochs), and benchmark --- completed in $\sim$4.5~hours unattended.

  \item[v5 (frozen Qwen weights):] Adding the frozen Qwen3.5 embedding table and LM head (1{,}017M params in BF16) to the drafter increased total memory to $\sim$3.2\,GB GPU. The 295M trainable drafter parameters trained for 10~epochs ($\sim$13\,min/epoch) reaching val\_loss=3.80 with 50.8\% per-token accuracy. However, this \emph{did not translate} to better $\tau$ because: (a)~multiple training runs crashed with OOM on checkpoint resume (VRAM exhausted at 15.47\,GB), (b)~702 extracted feature samples contained NaN/Inf values and had to be skipped, and (c)~the Qwen shared weights improved loss but not left-to-right acceptance.

  \item[v6 $\to$ v7 (RoPE regression):] Adding RoPE ($\theta{=}10^7$) alongside absolute positional information \emph{degraded} $\tau$ from 0.43 to 0.04 despite improving val\_loss from 5.06 to 4.02. The drafter receives position information implicitly through the target's hidden states; explicit RoPE created conflicting position signals that hurt sequential acceptance.

  \item[v8 (final config):] 8~layers with deep KV injection from 5~target layers (1, 10, 19, 28, 37), bidirectional GQA (32/4 heads), SwiGLU FFN (6{,}144 intermediate), likelihood loss ($\eta{=}3.0$). This is the architecture used in the offline benchmark. Three checkpoint variants were produced: offline (IQ3\_S features), online (c5 context), and online-IQ3 (quantized + context).
\end{description}

The critical insight is that \textbf{better validation loss does not predict better $\tau$}. v7 has the best val\_loss (4.02) among discrete versions but the worst $\tau$ (0.04). The v3 single-shot training paradigm with exponential loss is what drives acceptance, not raw next-token accuracy.

% ============================================================
\section{The GDN State Barrier}
\label{sec:barrier}

Despite $\tau{=}9.4$ offline, we were unable to deploy online speculative decoding on Qwen3.5-35B-A3B. We identify a fundamental structural incompatibility.

\subsection{The Problem}

In standard speculative decoding, when a draft token at position $i$ is rejected, the target model's KV cache is trimmed via \texttt{llama\_memory\_seq\_rm} to remove entries at positions $\geq i$. This works because KV caches are \emph{position-indexed}: each position's key-value pair is independent and can be selectively removed.

GDN recurrent state is fundamentally different:
\begin{equation}
  M_n = f(M_{n-1}, x_n) = \alpha_n \cdot M_{n-1} \cdot (I - k_n q_n^\top) + k_n v_n^\top
\end{equation}

The state $M_n$ is a \emph{lossy compression} of all tokens $x_{1:n}$. There is no operation to compute $M_{i-1}$ from $M_n$ --- the information about individual token contributions has been irreversibly compressed.

\subsection{Four Structural Walls}

We identify four obstacles in the \texttt{llama.cpp} runtime:

\begin{enumerate}
  \item \textbf{No \texttt{seq\_rm} for recurrent layers.} The \texttt{llama\_memory\_seq\_rm} function only operates on KV cache entries. Recurrent state has no equivalent trimming operation.
  \item \textbf{No state snapshot/restore API.} There is no mechanism to checkpoint $M_t$ before speculation and restore it upon rejection.
  \item \textbf{No batch verification with rollback.} The runtime cannot process $K$ draft tokens, then roll back the recurrent state to a specific position.
  \item \textbf{No separate state for draft vs.\ target.} Draft and target evaluations share the same recurrent state, so running the drafter corrupts the target's state.
\end{enumerate}

\subsection{Scope of Impact}

This barrier affects \emph{all} hybrid SSM-attention architectures:
\begin{itemize}
  \item Jamba~\citep{lieber2024jamba} (Mamba + attention)
  \item RWKV-6/7 (linear attention + channel mixing)
  \item Qwen3.5 MoE family (GDN + GQA)
  \item Any future model mixing recurrent and attention layers
\end{itemize}

Standard speculative decoding (EAGLE, Medusa, etc.) is structurally incompatible with these architectures unless the runtime provides explicit state management.

% ============================================================
\section{Workaround: \textsc{Chim\`ere-deltanet}}
\label{sec:workaround}

We developed \textsc{Chim\`ere-deltanet}, a Rust-native inference runtime (56K lines) with explicit GDN state management that bypasses the barrier by design:

\begin{enumerate}
  \item \textbf{State checkpoint}: before speculation, snapshot the GDN state $M_t$ for all 30 recurrent layers ($\sim$120\,MB at FP16).
  \item \textbf{Attention-only verification}: run batch verification only through the 10 attention layers (which have standard KV caches).
  \item \textbf{Selective restoration}: on rejection at position $i$, restore the GDN state from checkpoint and replay accepted tokens $1{:}i{-}1$ through the recurrent layers only ($\sim$2\,ms per token).
\end{enumerate}

This approach trades $\sim$120\,MB of memory and a small replay cost for correct speculative decoding on hybrid architectures. The runtime achieves 83\,tok/s on RTX~5060~Ti with full Engram memory integration.

Source code: \url{https://github.com/AIdevsmartdata/chimere}

% ============================================================
\section{Related Work}

\paragraph{Speculative decoding.}
\citet{leviathan2023fast} and \citet{chen2023accelerating} introduced speculative decoding. EAGLE~\citep{li2024eagle} and Medusa~\citep{cai2024medusa} propose learned drafters. EAGLE-3 extends to dynamic draft lengths. All assume standard KV cache architectures.

\paragraph{Block diffusion.}
BD3-LMs~\citep{arriaga2024block} introduced block diffusion for language modeling with block size interpolation. DFlash~\citep{dflash2024} applied it as a speculative decoding drafter.

\paragraph{SSM speculation.}
SpecMamba~\citep{zhang2024specmamba} identified challenges with speculative decoding for Mamba models but did not address hybrid SSM-attention architectures or provide a runtime solution.

% ============================================================
\section{Conclusion}

We demonstrated that block diffusion drafting can achieve $\tau{=}9.4$ on MoE targets (55.84\% acceptance over 72{,}908 blocks) with three architectural innovations and ${\sim}4{\times}$ less data than the original DFlash.
More importantly, we identified and formalized the \emph{GDN State Barrier} --- a fundamental incompatibility between speculative decoding and hybrid recurrent-attention architectures that affects a growing family of production models.
Our Rust runtime provides a practical workaround, and we release all code, data, and trained models to enable further research.

\paragraph{Limitations.}
Our offline $\tau$ measurement uses pre-extracted hidden states; online $\tau$ may differ due to distribution shift during autoregressive generation. The $\tau{=}9.4$ result uses iterative denoising (16 steps), which adds ${\sim}43\,\text{ms}$ of drafter compute per block ($16 \times 2.7\,\text{ms}$); single-shot inference gives $\tau{=}2.9$ at 2.7\,ms. The GDN state checkpoint approach increases memory by ${\sim}120\,\text{MB}$. Our evaluation uses a single model (Qwen3.5-35B-A3B) on a single GPU (RTX~5060~Ti). The DFlash original reports results on a different target (Qwen3-Coder-30B-A3B), so the comparison is indicative rather than controlled.

% ============================================================
\begin{thebibliography}{10}

\bibitem[Arriaga et~al.(2024)]{arriaga2024block}
Arriaga, L., et~al.
\newblock Block diffusion: Interpolating between autoregressive and diffusion language models.
\newblock \emph{arXiv preprint arXiv:2503.09573}, 2024.

\bibitem[Cai et~al.(2024)]{cai2024medusa}
Cai, T., et~al.
\newblock Medusa: Simple LLM inference acceleration framework with multiple decoding heads.
\newblock \emph{ICML}, 2024.

\bibitem[Chen et~al.(2023)]{chen2023accelerating}
Chen, C., et~al.
\newblock Accelerating large language model decoding with speculative sampling.
\newblock \emph{arXiv preprint arXiv:2302.01318}, 2023.

\bibitem[DFlash(2024)]{dflash2024}
z-lab.
\newblock DFlash: Accelerating autoregressive inference with block diffusion.
\newblock \emph{arXiv preprint arXiv:2602.06036}, 2024.

\bibitem[Leviathan et~al.(2023)]{leviathan2023fast}
Leviathan, Y., et~al.
\newblock Fast inference from transformers via speculative decoding.
\newblock \emph{ICML}, 2023.

\bibitem[Li et~al.(2024)]{li2024eagle}
Li, Y., et~al.
\newblock EAGLE: Speculative sampling requires rethinking feature uncertainty.
\newblock \emph{ICML}, 2024.

\bibitem[Lieber et~al.(2024)]{lieber2024jamba}
Lieber, O., et~al.
\newblock Jamba: A hybrid transformer-mamba language model.
\newblock \emph{arXiv preprint arXiv:2403.19887}, 2024.

\bibitem[Yang et~al.(2024)]{yang2024gated}
Yang, S., et~al.
\newblock Gated delta networks: Improving mamba-2 with delta rule.
\newblock \emph{ICLR}, 2025.

\bibitem[Zhang et~al.(2024)]{zhang2024specmamba}
Zhang, P., et~al.
\newblock SpecMamba: Speculative decoding for Mamba architectures.
\newblock \emph{arXiv preprint}, 2024.

\end{thebibliography}

\end{document}
